---
title: "Generalized additive models"
author: "ST310"
date: "3/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(broom)
library(modelr)
library(gam)
theme_set(theme_minimal(base_size = 20))
```

## Simulate data from an additive model

Change the functions, add other predictors with their own functions which could be linear or nonlinear, change the sample size or the `sd` of the `rnorm` noise (or the distribution of the noise if you're interested in robustness)

```{r}
f1 <- function(x) -1 + 2*x + x^2 
f2 <- function(x) sin(pi*x) 
f3 <- function(x) exp(-5*(x-1)^2)

set.seed(1)
n <- 1000
confounder = rnorm(n, sd = .5)
df <- data.frame(
  x1 = 2*(runif(n)-1/2) + confounder,
  x2 = sample(1:100 / 50,  n, replace = TRUE),
  x3 = 2*rbeta(n, 1, 3) + confounder
) %>% 
  mutate(
    y = f1(x1) + f2(x2) + f3(x3) + rnorm(n, sd = 1)
)
```

Fit a GAM, show the partial dependence plots

```{r}
fit_gam <- gam(y ~ lo(x1) + lo(x2) + lo(x3), data = df)
```

Try `plot`, and if you have the `plotmo` library try `plotmo` with `all2 = TRUE`

```{r}
summary(fit_gam)
```

Don't need to change anything in this code chunk

```{r}
# predict for range of xj values
partial_predict <- function(j, df, obj) {
  xj <- paste0("x", j)
  newdf <- df %>% 
    mutate(across(starts_with("x") & -xj, mean))
  predict(obj, newdata = newdf)
}

# plot y vs xj
partial_plot <- function(j) {
  xj <- paste0("x", j)
  fj <- paste0("f", j)
  ggplot(df, aes(get(xj), y)) +
    geom_point(alpha = .5) +
    geom_smooth() +
    geom_line(aes(y = partial_predict(j, df, fit_gam)), color = "red") + 
    xlab(xj) + ylab("") +
    geom_function(fun = get(fj),
                  size = 1,
                  color = "green") +
    theme(axis.text.x=element_blank(),
          axis.ticks.x=element_blank())
}
```

Might want to change below if you add more predictors

```{r}
p1 <- partial_plot(1) + ylab("y")
p2 <- partial_plot(2)
p3 <- partial_plot(3)
```

Side by side plots by adding with the `patchwork` library

```{r message = FALSE}
library(patchwork)
p1 + p2 + p3
```



Why might some of the estimated functions be shifted up/down?

## Breaking assumptions

Try generating data that does not satisfy the additivity assumption

```{r}
f1 <- function(x) -1 + 2*x + x^2 
f2 <- function(x) sin(pi*x) 
f3 <- function(x) exp(-5*(x-1)^2)

set.seed(1)
n <- 1000
confounder = rnorm(n, sd = .5)
df <- data.frame(
  x1 = 2*(runif(n)-1/2) + confounder,
  x2 = sample(1:100 / 50,  n, replace = TRUE),
  x3 = 2*rbeta(n, 1, 3) + confounder
) %>% 
  mutate(
    y = x1 * x2^2  + f1(x1) + f2(x2) + f3(x3) + rnorm(n, sd = 1)
)
```

Fit a GAM, show the partial dependence plots

```{r}
fit_gam <- gam(y ~ lo(x1) + lo(x2) + lo(x3), data = df)
```


```{r}
p1 <- partial_plot(1) + ylab("y")
p2 <- partial_plot(2)
p3 <- partial_plot(3)
p1 + p2 + p3
```


## Sparse additive non-linear models

Generate more predictors which are just noise

Use the `gamsel` package to select which variables to include, and which variables are modeled linearly vs non-linearly

```{r message = FALSE}
library(gamsel) # install.packages("gamsel")
```


```{r}
n <- nrow(df)
high_dim_df <- df %>% mutate(
  x4 = rnorm(n) + confounder,
  x5 = runif(n),
  x6 = rbeta(n, 2, 1)
) 
```




```{r}
fit_gam <- gam(y ~ lo(x1) + lo(x2) + lo(x3) +
                 lo(x4) + lo(x5) + lo(x6),
               data = high_dim_df)
```


```{r}

partial_plot <- function(j) {
  xj <- paste0("x", j)
  fj <- paste0("f", j)
  ggplot(high_dim_df, aes(get(xj), y)) +
    geom_point(alpha = .5) +
    geom_smooth() +
    geom_line(aes(y = partial_predict(j, high_dim_df, fit_gam)), color = "red") + 
    xlab(xj) + ylab("") +
    geom_function(fun = get(fj),
                  size = 1,
                  color = "green") +
    theme(axis.text.x=element_blank(),
          axis.ticks.x=element_blank())
}

f4 <- function(x) 0
f5 <- function(x) 0
f6 <- function(x) 0
p1 <- partial_plot(1) + ylab("y")
p2 <- partial_plot(2)
p3 <- partial_plot(3)
p4 <- partial_plot(4) + ylab("y")
p5 <- partial_plot(5)
p6 <- partial_plot(6)
p1 + p2 + p3
```

```{r}
p4 + p5 + p6
```

```{r}
X <- high_dim_df %>% select(-y)
Y <- high_dim_df$y
fit_gamsel <- gamsel(X, Y)
```


```{r}
summary(fit_gamsel)
```


```{r}
cv_fit_gamsel <- cv.gamsel(X, Y)
```


```{r}
cv_fit_gamsel$lambda.min
```

```{r}
plot(fit_gamsel, newx = X)
```


## Test several models using tidymodels

```{r}
library(tidymodels)

```

