---
title: "Week N seminar: <topic>"
author: "ST310"
format:
  html:
    toc: true
    embed-resources: false
execute:
  echo: true
  warning: false
  message: false
  error: false
---

<!--
TEMPLATE (delete in a real notebook). Student version. Rules:
  - About 25 minutes of problems (mostly mathematical, some conceptual, from the
    item bank, with light scaffolding; the teacher reveals hints after enough
    time), then the rest coding. Minutes go in HTML comments, not student text.
  - Blanks are `...`. The generator gives every chunk that contains a blank
    `#| eval: false`, so the student notebook renders with the blanks unfilled
    and no error blocks; a chunk that only *uses* an object defined in a
    blanked chunk needs `#| eval: !expr exists("that_object")` in the complete
    file. `error: false` in both files, so a broken solution fails the
    instructor render rather than rendering an error block.
  - Minutes go in HTML comments (`<!-- 25 min -->`), never in headings or
    student text. Scaffold the boilerplate (a plot skeleton, a loop skeleton,
    a function signature) as a commented line above the blank; the student
    writes the statistical operation.
  - Hints go in `<!-- HINT: ... -->` and come out collapsed; a checkpoint chunk
    (`stopifnot(...)`, guarded with `eval: !expr exists(...)`) after each
    stage of a long coding block lets a stuck student see which stage to fix.
  - Code is minimal: no RNGkind, no set.seed unless it is essential (students
    should see how much changes with a different split), no `;`, but keep the
    theme settings and other aesthetics.
  - Retrieval opener first, exit task last.
  - Default dataset first, alternatives at the end with a loader chunk each and
    a link to the data note in data/.
  - Attrition data, wherever used, are labelled fictional in the student text.
  - Never move answers from the _complete version into this file; the student
    file is generated from the complete one by _build_env/make_incomplete.py.
-->

```{r}
#| label: setup
library(tidyverse)
theme_set(theme_minimal())
```

## Before you start {.retrieval}

<!-- 5 min -->

Without looking anything up, ...

## Problems

<!-- 25 min -->

**1.** ...

**2.** ...

## Coding

<!-- 55 min -->

### <segment>

```{r}
# ...
```

**Question.** ...

## Before you leave {.retrieval}

<!-- 5 min -->

In two sentences: ...

## Alternative datasets

If you would rather work with different data, each loader below produces the same columns the code above expects. See the data notes for provenance and licence.

```{r}
#| eval: false
# alternative 1 (data/<name>.md)
```
