---
title: "Non-linear and local methods"
author: "Joshua Loftus"
date: "2/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) # required
library(broom)     # required
library(modelr)    # for data_grid
theme_set(theme_minimal(base_size = 22))
```

## 2d non-linear classification data

```{r}
# sample size
n <- 800
# function with zero set that
# defines the perfect (Bayes) decision boundary
true_boundary_function <- function(x1, x2) {
  # experiment with changing this
  thing <- x1^2*x2^3
  (x1^2 + x2^2 - 1)^3 - thing
}
train <- data.frame(
  # change the distribution
  # or scale as needed
  x1 = 1.5*(1 - 2*runif(n)),
  x2 = 1.5*(1 - 2*runif(n))
) %>% 
  mutate(
    # labels if classes were perfectly separated
    separable = true_boundary_function(x1,x2) > 0,
    # labels if classes are "noisy"
    y = factor(rbinom(n, 1, 9/10 - (8/10) * as.numeric(separable)))
  )
```

### Plot the data

```{r}
train_plot <-
  ggplot(train, aes(x1, x2)) +
  geom_point(aes(shape = y, color = y),
             alpha = .5, show.legend = FALSE) +
  xlab("") + ylab("")
  
train_plot
```

### Plot the Bayes decision boundary

```{r}
decision_surface <- 
  data_grid(train,
          x1 = seq_range(x1, 300, expand = .05),
          x2 = seq_range(x2, 300, expand = .05)) %>%
  mutate(z = true_boundary_function(x1, x2))

boundary_plot <-
  train_plot +
  geom_contour(
    data = decision_surface,
    aes(x1, x2, z=z),
    bins = 2,
    size = 1,
    color = "black",
    alpha = .5)

boundary_plot
```

### How well does linear classification do?

Also try changing the formula to fit a logistic regression model with non-linear transformations of the predictors

```{r}
# Fit model
logit_model <-  
  glm(y ~ x1 + x2 + poly(x1,2) * poly(x2,2), family = "binomial", data = train)
# try formula + poly(x1,2) * poly(x2,2)

# Generate decision boundary
logit_surface <- logit_model %>%
   augment(type.predict = "response",
              newdata = decision_surface)
```

Plot decision boundary of logistic model

```{r}
boundary_plot +
  geom_contour(
    data = logit_surface,
    aes(x1, x2, z = .fitted),
    size = 1,
    color = "green",
    bins = 2)
```

```{r}
summary(logit_model)
```

## 1-d smooth regression example

Generate a one-dimensional example with a non-linear relationship

```{r}
# change this function
f <- function(x) sin(4*pi*x)
n <- 200
train1d <- 
  data.frame(
    x = rbeta(n, 1, 3)
    ) %>%
  # change the noise level sd
  mutate(y = f(x) + rnorm(n, sd = .1))

# plot data and loess curve
ggplot(train1d, aes(x, y)) + 
  geom_point() +
  geom_smooth()
```

Linear regression with a polynomial transform of x

```{r}
model_lm <- lm(y ~ poly(x, 5), data = train1d)

train1d_grid <- 
  data_grid(train1d,
          x = seq_range(c(x, .78), 500, expand = .05))

augment(model_lm,
        newdata = train1d_grid) %>%
  ggplot(aes(x, y)) +
  geom_point(data = train1d) +
  geom_line(aes(y = .fitted))
```


## Local methods


### Local averages/regression for 1d example

Write functions to compute:

- Average of k-nearest neighbors

- Local regression:

1. Gather the fraction s = k/n of training points whose xi are closest
to x0.
2. Assign a weight Ki0 = K(xi,x0) to each point in this neighborhood, so that the point furthest from x0 has weight zero, and the closest has the highest weight. All but these k nearest neighbors get weight zero.
3. Fit a weighted least squares regression of the yi on the xi using the aforementioned weights
4. Use the weighted least squares model to predict for x0

```{r}
# assume input data has variables x and y
# assume data is sorted on x from smallest to largest
local_regression <- function(data, x0, s = .1) {
  n <- nrow(data)
  k <- ceiling(n*s)
  i <- 0
  loop <- TRUE
  while (loop) {
    i <- i + 1
    if (data$x[i] > x0) {
      loop <- FALSE
    }
    if (i >= n) {
      loop <- FALSE
    }
  }
  # i such that
  # data$x[i] > x0
  # but data$x[i-1] <= x0
  local_range <- (i - ceiling(k/2)) : (i + floor(k/2))
  
  if (min(local_range) < 1) {
    local_range <- 1:k
  } else if (max(local_range) > n) {
    local_range <- (n-k):n
  }
  
  local_lm <- lm(y ~ x, data = data[local_range, ])
  sum(coef(local_lm) * c(1, x0))
}
```

Generate predictions from these methods and plot them

```{r}
train_local <- train1d %>%
  arrange(x)

local_regression(train_local, x0 = .2)
```


Plot

```{r}
for (i in 1:nrow(train1d_grid)) {
  train1d_grid$.fitted[i] <- local_regression(train_local, x0 = train1d_grid$x[i], s = .05)
}

  ggplot(train1d_grid, aes(x, y)) +
  geom_point(data = train1d) +
  geom_line(aes(y = .fitted))
```

### Classify 2d example with k-nearest neighbors

Write a function to implement k-NN classification

You may use matrix operations but not any more sophisticated built-in R functions (or libraries)

Hint: https://en.wikipedia.org/wiki/Euclidean_distance_matrix#Relation_to_Gram_matrix

```{r}

```

Plot the data and decision boundary for your k-NN classifier

```{r}

```


