---
title: "Non-linear and local methods"
author: "Joshua Loftus"
date: "2/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) # required
library(broom)     # required
library(modelr)    # for data_grid
theme_set(theme_minimal(base_size = 22))
```

## 2d non-linear classification data

```{r}
# sample size
n <- 800
# function with zero set that
# defines the perfect (Bayes) decision boundary
true_boundary_function <- function(x1, x2) {
  # experiment with changing this
  abs(x1-.5) - abs(2*x2+.5) + .3
}
train <- data.frame(
  # change the distribution
  # or scale as needed
  x1 = 1.5*(1 - 2*runif(n)),
  x2 = 1.5*(1 - 2*runif(n))
) %>% 
  mutate(
    # labels if classes were perfectly separated
    separable = true_boundary_function(x1,x2) > 0,
    # labels if classes are "noisy"
    y = factor(rbinom(n, 1, 9/10 - (8/10) * as.numeric(separable)))
  )
```

### Plot the data

```{r}
train_plot <-
  ggplot(train, aes(x1, x2)) +
  geom_point(aes(shape = y, color = y),
             alpha = .5, show.legend = FALSE) +
  xlab("") + ylab("")
  
train_plot
```

### Plot the Bayes decision boundary

```{r}
decision_surface <- 
  data_grid(train,
          x1 = seq_range(x1, 300, expand = .05),
          x2 = seq_range(x2, 300, expand = .05)) %>%
  mutate(z = true_boundary_function(x1, x2))

boundary_plot <-
  train_plot +
  geom_contour(
    data = decision_surface,
    aes(x1, x2, z=z),
    bins = 2,
    size = 1,
    color = "black",
    alpha = .5)

boundary_plot
```

### How well does linear classification do?

Also try changing the formula to fit a logistic regression model with non-linear transformations of the predictors

```{r}
# Fit model
logit_model <-  
  glm(y ~ x1 + x2, family = "binomial", data = train)
# try formula + poly(x1,2) * poly(x2,2)

# Generate decision boundary
logit_surface <- logit_model %>%
   augment(type.predict = "response",
              newdata = decision_surface)
```

Plot decision boundary of logistic model

```{r}
boundary_plot +
  geom_contour(
    data = logit_surface,
    aes(x1, x2, z = .fitted),
    size = 1,
    color = "green",
    bins = 2)
```

```{r}
summary(logit_model)
```

## 1-d smooth regression example

Generate a one-dimensional example with a non-linear relationship

```{r}
# change this function
f <- function(x) x * sin(1/x^2)
n <- 50
train1d <- 
  data.frame(
    x = rbeta(n, 1, 3)
    ) %>%
  # change the noise level sd
  mutate(y = f(x) + rnorm(n, sd = .1))

# plot data and loess curve
ggplot(train1d, aes(x, y)) + 
  geom_point() +
  geom_smooth()
```

Linear regression with a polynomial transform of x

```{r}

```


## Local methods


### Local averages/regression for 1d example

Write functions to compute:

- Average of k-nearest neighbors

- Local regression:

1. Gather the fraction s = k/n of training points whose xi are closest
to x0.
2. Assign a weight Ki0 = K(xi,x0) to each point in this neighborhood, so that the point furthest from x0 has weight zero, and the closest has the highest weight. All but these k nearest neighbors get weight zero.
3. Fit a weighted least squares regression of the yi on the xi using the aforementioned weights
4. Use the weighted least squares model to predict for x0

```{r}

```

Generate predictions from these methods and plot them

```{r}

```

### Classify 2d example with k-nearest neighbors

Write a function to implement k-NN classification

You may use matrix operations but not any more sophisticated built-in R functions (or libraries)

Hint: https://en.wikipedia.org/wiki/Euclidean_distance_matrix#Relation_to_Gram_matrix

```{r}

```

Plot the data and decision boundary for your k-NN classifier

```{r}

```


