---
title: "Nonlinearity and nearest neighbours"
author: "Joshua Loftus"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) # required
library(broom)     # required
library(modelr)    # for data_grid
theme_set(theme_minimal(base_size = 20))
```

## 2d non-linear classification data

```{r}
# sample size
n <- 800
# function with zero set that
# defines the perfect (Bayes) decision boundary
true_boundary_function <- function(x1, x2) {
  # experiment with changing this
  thing <- x1^2*x2^3
  (x1^2 + x2^2 - 1)^3 - thing^2
}
train <- data.frame(
  # change the distribution
  # or scale as needed
  x1 = 1.5*(1 - 2*runif(n)),
  x2 = 1.5*(1 - 2*runif(n))
) %>% 
  mutate(
    # labels if classes were perfectly separated
    separable = true_boundary_function(x1,x2) > 0,
    # labels if classes are "noisy"
    y = factor(rbinom(n, 1, 9/10 - (8/10) * as.numeric(separable)))
  )
```

### Plot the data

```{r}
train_plot <-
  ggplot(train, aes(x1, x2)) +
  geom_point(aes(shape = y, color = y),
             alpha = .5, show.legend = FALSE) +
  xlab("") + ylab("")
  
train_plot
```

### Plot the Bayes decision boundary

```{r}
decision_surface <- 
  data_grid(train,
          x1 = seq_range(x1, 300, expand = .05),
          x2 = seq_range(x2, 300, expand = .05)) %>%
  mutate(z = true_boundary_function(x1, x2))

boundary_plot <-
  train_plot +
  geom_contour(
    data = decision_surface,
    aes(x1, x2, z=z),
    bins = 2,
    size = 1,
    color = "black",
    alpha = .5)

boundary_plot
```

### How well does linear classification do?

Also try changing the formula to fit a logistic regression model with non-linear transformations of the predictors

```{r}
# Fit model
logit_model <-  
  glm(y ~ x1 + x2 + poly(x1,2) * poly(x2,2), family = "binomial", data = train)
# try formula + poly(x1,2) * poly(x2,2)

# Generate decision boundary
logit_surface <- logit_model %>%
   augment(type.predict = "response",
              newdata = decision_surface)
```

Plot decision boundary of logistic model

```{r}
boundary_plot +
  geom_contour(
    data = logit_surface,
    aes(x1, x2, z = .fitted),
    size = 1,
    color = "green",
    bins = 2) 
```

### Classify 2d example with k-nearest neighbors

Write a function to implement k-NN classification

You may use matrix operations and simple functions like `sort` but not any more sophisticated built-in R functions (or libraries)

#### Write a function that computes some distance between two points

You can choose which notion of distance you want to use

```{r}
# Euclidean or L2 distance
Dist <- function(x1, x2) {
  sum((x1-x2)^2) # sqrt doesn't change which points are nearest
}
```

#### Write a function that outputs the y values for the k-nearest nearest neighbors of a point x0

```{r}
# Requires a distance function D
# to already be defined
nearest_y_values <- function(x0, k, x, y) {
  n <- nrow(x)
  distances <- rep(0, n)
  for (i in 1:n) {
    distances[i] <- Dist(x0, x[i,])
  }
  k_neighborhood <- order(distances, decreasing = FALSE)[1:k]
  y[k_neighborhood]
}
```

#### Write a function that outputs a classification using majority vote

```{r}
knn <- function(x0, k, x, y) {
  votes <- nearest_y_values(x0, k, x, y)
  levels(y)[which.max(table(votes))]
}
```


#### Evaluate your function at various points, using various values of k, and compare the output to the plotted training data

Try points where the nearest neighbors are clearly majority 0 or 1, and a point where the portion is relatively even. Run the code below a few times and see if the answers change, then pick a different `K` and try again

```{r}
K <- 1
bootstrap_sample <- sample(1:nrow(train), replace = TRUE)
x <- train[bootstrap_sample, 1:2]
y <- train$y[bootstrap_sample]
```


In a majority 1 region

```{r}
knn(x0 = c(0,0), k = K, x, y)
```

In a majority 0 region

```{r}
knn(x0 = c(0,1.2), k = K, x, y)
```

In a relatively even region

```{r}
knn(x0 = c(1,1), k = K, x, y)
```
