---
title: "Interpretable ML package"
author: "Joshua Loftus"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)  
library(iml)
library(palmerpenguins)  
library(randomForest)
library(tree)
library(gam)
theme_set(theme_minimal(base_size = 22))
```

## Random forest to classify penguins

```{r}
pg <- penguins %>%
  # not interested in classifying by time/island
  dplyr::select(-island, -year, -sex) %>%
  drop_na()
```

#### First fit a single tree

```{r}
fit_tree <- tree(species ~ .,
                 split = "gini",
                 control = tree.control(nrow(pg), mincut = 40),
                 data = pg)
fit_tree
#plot(fit_tree, type = "uniform")
#text(fit_tree, pretty = 0, cex = 1.1)
```

#### Experiment with changing the control parameters

#### Fit a random forest

```{r}
fit_rf <- randomForest(species ~ ., ntree = 100, mtry = 2,
                  minsize = 20, data = pg)
fit_rf
```

#### Variable importance plot

Read `?importance` to learn more

```{r}
varImpPlot(fit_rf)
```

#### Experiment with changing the control parameters

#### Partial dependence plots

```{r}
pred_rf <- Predictor$new(fit_rf)
pdp_rf <- FeatureEffects$new(pred_rf,
      features = c("bill_length_mm",
                   "flipper_length_mm"),
      method = "pdp+ice")
plot(pdp_rf) 
```

#### Check other variables

#### Interaction plots

```{r}
int_rf <- Interaction$new(pred_rf)
plot(int_rf)
```

#### For each variable, how much does the model's predictions depend on the interaction of that variable with others or by the variable alone?

Hint: read `?Interaction`

If the measure is close to 0 then the predictions depend mostly on the variable alone, if it's close to 1 then the predictions depend more on interactions between that variable and other predictors

#### Fit a gam and interpret the interaction plot for the fitted model

Note that the `iml` and `gam` packages are slightly less compatible, e.g. it seems that `iml` plots will automatically show results for only 1 class

```{r}
fit_gam <- gam(species ~ ., family = "binomial",  data = pg)
fit_gam
```

```{r}
pred_gam <- Predictor$new(fit_gam)#, type = "response")
pdp_gam <- FeatureEffects$new(pred_gam,
      features = c("bill_length_mm",
                   "flipper_length_mm"),
      method = "pdp+ice")
plot(pdp_gam) 
```

```{r}
int_gam <- Interaction$new(pred_gam)
plot(int_gam) + xlim(c(0,1))
```

#### Is this what we expect?

#### Note that the additivity of the model is on the log-odds scale. Change the prediction type to "response" and check the interaction plot again
