---
title: "Gradient descent"
author: "Joshua Loftus"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
theme_set(theme_minimal(base_size = 12))
```

## Generating simulated data

```{r}
# Setting parameters
set.seed(1)
# Try changing some parameters
n <- 200
p <- 5
sparsity <- 2
nonzero_beta <- c(1, -1)
# or, e.g. rep(1, sparsity), runif(sparsity, min = -1, max = 1)
true_beta <- c(.5, nonzero_beta, rep(0, p - sparsity))
# Note: we set the intercept to 0
```

```{r}
# Generating simulated data
X <- cbind(rep(1, n), matrix(rnorm(n*p), nrow = n))
mu <- X %*% true_beta
px <- exp(mu)/(1+exp(mu))
Y <- rbinom(n, 1, prob = px)
train_ld <- data.frame(y = as.factor(Y), x = X)
fit_glm <- glm(Y ~ X-1, family = "binomial")
```

## Log-likelihood function

```{r}
# See e.g. lecture slides 
logL <- function(X, Y, beta) {
  Xbeta <- X %*% beta
  expXbeta <- exp(-Xbeta)
  exp_ratio <- 1/(1+expXbeta)
  sum(Y * log(exp_ratio) + (1-Y) * log(1-exp_ratio))
}
```

## Numeric differentiation

Reference: [Wikipedia](https://en.wikipedia.org/wiki/Numerical_differentiation)

```{r}
# Note: the step size h here is chosen to be
# on the order of magnitude of the cube-root
# of .Machine$double.eps, which helps avoid
# some numerical approximation errors
numeric_grad <- function(X, Y, beta, h = 1e-06) {
  numerator <- sapply(1:length(beta), function(j) {
    H <- rep(0, length(beta))
    H[j] <- h # step in coordinate j
    logL(X, Y, beta + H) - logL(X, Y, beta - H)
  })
  numerator / (2*h)
}
```

## Gradient descent

Reference: [Wikipedia](https://en.wikipedia.org/wiki/Gradient_descent)

Note: Below we use the *Barzilai–Borwein method* for choosing a step size. If we did not use an *adaptive* method like this then we would have to experiment to find something that worked, and most likely we would need to redo that trial and error each time we tried a different problem

```{r}
# Initialize and take first step
beta_prev2 <- rnorm(ncol(X)) 
grad_prev2 <- numeric_grad(X, Y, beta_prev2)
beta_prev1 <- beta_prev2 + 0.1 * grad_prev2 / sqrt(sum(grad_prev2^2))
grad_prev1 <- numeric_grad(X, Y, beta_prev1)
previous_loss <- logL(X, Y, beta_prev2)
next_loss <- logL(X, Y, beta_prev1)
beta_mat <- rbind(beta_prev2,
                  beta_prev1,
                  matrix(0, nrow = 50, ncol = ncol(X)))
rownames(beta_mat) <- NULL
beta_path <- data.frame(step = 0:51, b = beta_mat)
steps <- 1
```


```{r}
# Repeat until convergence
tol <- 1e-5 # (error) tolerance
while (abs(previous_loss - next_loss) > tol) {
  grad_diff <- grad_prev1 - grad_prev2
  step_BB <- sum((beta_prev1 - beta_prev2) * grad_diff) / sum(grad_diff^2)
  
  beta_prev2 <- beta_prev1
  beta_prev1 <- beta_prev1 - step_BB * grad_prev1
  
  grad_prev2 <- grad_prev1
  grad_prev1 <- numeric_grad(X, Y, beta_prev1)  
  
  previous_loss <- next_loss
  next_loss <- logL(X, Y, beta_prev1)
  
  if (round(steps/5) == steps/5) print(previous_loss)
  steps <- steps + 1
  beta_path[steps, 2:ncol(beta_path)] <- beta_prev1
}
```

How many steps?

```{r}
steps
```

#### Did it work?

```{r}
rbind(
  fit_glm |> coef(),
  beta_prev1,
  true_beta
)
```

Coefficient paths

```{r}
beta_path <- beta_path[1:steps, ]
beta_path |>
  pivot_longer(cols = starts_with("b"),
               names_to = "variable",
               values_to = "beta_hat") |>
  ggplot(aes(x = step, y = beta_hat, color = variable)) +
  geom_point() + geom_line() #+ scale_color_viridis_d()
```

(Note: the coefficient indexes may be a bit confusing because of the intercept term)

## How cool is this, exactly?

**Note**: with numeric differentiation we did not need to do the calculus ourselves to find an expression for the gradient. (People have done a lot of work clearing this path for us, though...)

**Question**: Suppose that instead of logistic regression we wanted to fit some other kind of model. What would we need to change about the code?

**Answer**: Only the `logL` function! 
