---
title: "Week 1 seminar"
author: "Joshua Loftus"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gapminder)
library(broom)
theme_set(theme_minimal()) # or whatever you prefer
```

## Repeating the gapminder analysis

Let's start easy by just repeating some steps but with data from a different year.

### Create the `gapminder` scatterplot using data from the year 2002




#### Create a simple lm model to predict lifeExp

```{r}

```

#### Create a polynomial regression model to predict lifeExp

```{r}

```

#### Plot showing the two models

```{r}

```

Mean-square error of both models

```{r}
mean(residuals(model_lm)^2)
mean(residuals(model_poly)^2)
```

## Predicting on new data

Models are supposed to capture/use structure in the data that corresponds to structure in the real world. And if the real world isn't misbehaving, that structure should be somewhat stable.

For example, suppose the relationship changed dramatically from one time period to another time period. Then it would be less useful/interesting to have a model fit on data at one time period, because the same model might have a poor fit on data from a different time period.

Let's explore this with our `gapminder` models

#### Predictions on different years

Create datasets for the desired years

```{r}

```

Predict using `newdata` argument, then `pull` the residuals from the resulting data.frame

```{r}

```

#### Check 1997

```{r}
mean(lm_resid1997^2)
mean(loess_resid1997^2)
```


#### Check 2007

```{r}

```



## Conclusion/notes




